A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-06 Simulation/Writeup/NoteConvergence.tex
Update pdf to review
Feedback Gerko on Introduction.

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which can yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e. imputed) several times. The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \proglang{R} package \pkg{mice} \citep{mice}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods. More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}. All programming code used in this note is available on Github repository \href{https://github.com/gerkovink/shinyMice}{`ShinyMICE'}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the 'original' notation by \cite{rubin87}, are described in \citep[\S~XYZ]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably. Both refer to $Y_{mis}$, e.g. `$NA$' values in \proglang{R} or `999' in \proglang{SPSS}. The terms 'incomplete' or 'observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=of mids] { };
\node[main] (mids3) [below=of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=of mira] { };
\node[main] (mira3) [below=of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of main steps in multiple imputation (Adapted from \cite[\S~1.4.1]{buur18})}
\end{figure}

@@

This note is primarily concerned with the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain` of potential values is considered. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling). The collection of values (or summary statistics) across iterations of one of the $m$ imputations will be referred to as `imputation chain'.

\textbf{ADD THIS? The process of deliberately creating missingness in an otherwise complete data set is referred to as `amputing' the data.}

\textbf{[explain difference MI, FCS, mice]}

@@

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{steps.png}
%   }
%   \caption{Scheme of main steps in multiple imputation (Adapted from \cite[par.~1.4.1]{buur18})}
%     \label{resultssim}
% \end{figure}

% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

@@

%% -- Theoretical Background ---------------------------------------------------------------
\subsection{Theoretical Background (NOG NIET AF VANAF HIER)} \label{sec:background}

@@

@@

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.

>>>>

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `imputed' (i.e. filled in) several times. The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods. More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably. Both refer to $Y_{mis}$, e.g. `\texttt{NA}' values in \proglang{R} or `\texttt{999}' in \proglang{SPSS}. The terms `incomplete' or `observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \citep[\S~1.4.1]{buur18}.}
\end{figure}

@@

This note focusses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is considered. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling). The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.

@@

% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

@@

%% -- Theoretical Background ---------------------------------------------------------------
\subsection{Theoretical Background} \label{sec:background}

@@

High R hat would indicate .... low R hat values indicate ....

The second component of convergence ... interpreted as stability over iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.

@@

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.

\textbf{add loess line to plots to see when it becomes a flat line.}
